---
title: "Donor Responses to Civil Society Repression"
subtitle: "Chapter for the *Handbook of Political Control*"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
bibliography: references.bib
csl: csl/american-sociological-association.csl
toc: true
format:
  wordcount-html: 
    wordcount-banner: true
params:
  wordcount: |
    <strong>{{< words-sum body-note-ref >}} total words</strong>: {{< words-body >}} in the body • {{< words-ref >}} in the references • {{< words-note >}} in the notes
---

::: {.content-visible when-meta="wordcount-banner"}
{{< include _extensions/andrewheiss/wordcount/banner.html >}}
:::

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE, include = FALSE, warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = (6 * 0.618), fig.retina = 3,
  out.width = "100%", fig.align = "center"
)
```

```{r}
#| label: libraries-data
library(tidyverse)
library(tidygraph)
library(ggraph)
```

Given the exponential increase in administrative and legal repression of CSOs across the globe and its effectiveness at muzzling groups, how have donors responded? How do these responses differ across foreign government and private donors? How have local philanthropists and NGOs adapted and responded to such repression? What challenges do civil society groups continue to face despite these responses? 

## Foreign aid donors

A rich literature on the purposes of foreign aid explores a broad collection of reasons why states transfer money and resources to others. Aid agencies and international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with. Donor agencies rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011; @Kallman:2020]. In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy–oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017; @Schuller:2012], including being responsive to popular democratic pressure [@HeinrichKobayashiLong:2018], insulating themselves from regional instability [@Bermeo:2017], expanding export markets [@BarthelNeumayerNunnenkamp:2014; @Mitchell:2002], influencing recipient states' domestic policies [@Apodaca:2013], and projecting donor country values [@MorgensternBrown:2022].

### Donor responses and conditionality

A key enforcement mechanism to ensure that recipient-country actions align with donor-country goals—and thus reduce the moral hazard inherent in donating resources—is to condition aid on the achievement of specific democratizing and economic goals [@Winters:2024; @CheesemanSwedlundOBrien-Udry:2024; @HoefflerJustino:2024; @Kilby:2024; @Koch:2015]. Many donor governments condition aid on human rights outcomes, using progress in respect for human rights as leverage to shape recipient-country behavior. Conditional aid sends dual signals: it serves as an instrumental "stick" to pressure governments to change policies and an expressive "flag" to signal donor commitment to human rights norms [@DasandiErez:2023]. However, this dual-purpose strategy often creates dilemmas—if aid is withdrawn, it may jeopardize humanitarian programs or worsen political conditions. Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor’s human rights agenda. 

The often-conflicting trends of both pursuing normative development goals and furthering foreign policy interests leads to mixed responses when recipient countries fail to adhere to aid conditions and violate human rights. Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that when confronted by recipient states that violate human rights, donor agencies punish repressive states by reducing, suspending, or withdrawing aid [@LebovicVoeten:2009], increasing aid (or withdrawing less aid) to repressive states as a way to counter state violence [@Demirel-PeggMoskowitz:2009; @Nielsen:2013; @HeinrichKobayashiLong:2018; @EsareyDeMeritt:2017], or make no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b]. 

One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid. In reality, however, the decision is far more complex and involves a wide range of possible motivations and responses. @CheesemanSwedlundOBrien-Udry:2024 offer a stylized decision tree of possible choices following recipient-country human rights violations—we provide a simplified version of their framework in @fig-agency-response-tree. 

```{r}
#| label: fig-agency-response-tree
#| fig-cap: Possible donor- and recipient-country responses to human rights violations
#| include: true
#| fig-width: 6
#| fig-height: 3.8
#| out-width: 80%

nodes <- tribble(
  ~id, ~x, ~y,  ~step,
  1,   0,  2.5, "Violate human rights",
  2,   1,  4,   "Do nothing",
  3,   1,  3,   "Threaten",
  4,   1,  2,   "Withdraw\nor suspend",
  5,   1,  1,   "Shift aid",
  6,   2,  4,   "Do nothing",
  7,   2,  2.5, "Modify to improve\nrelationship",
  8,   2,  1,   "Pressure donors\nto back down"
)

edges <- tibble(
  from = c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5),
  to =   c(2, 3, 4, 5, 6, 7, 8, 6, 7, 8, 6, 7, 8, 6, 7, 8)
)

actors <- tribble(
  ~actor, ~label, ~x,
  "Recipient", "Recipient government\naction", 0,
  "Donor", "Donor government\nresponse", 1,
  "Recipient", "Recipient government\nresponse", 2
) |>
  mutate(
    xmin = x - 0.5,
    xmax = x + 0.5
  )

tidy_graph <- tbl_graph(nodes = nodes, edges = edges) |> 
  create_layout(layout = nodes)

ggraph(tidy_graph) +
  geom_rect(
    data = actors, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = 4.2, fill = actor), inherit.aes = FALSE,
    alpha = 0.25
  ) +
  geom_rect(
    data = actors, aes(xmin = xmin, xmax = xmax, ymin = 4.2, ymax = Inf, fill = actor), inherit.aes = FALSE,
    alpha = 0.5
  ) +
  geom_text(
    data = actors, aes(x = x, y = 4.4, label = label), inherit.aes = FALSE,
    size = 3.5,
    fontface = "bold", lineheight = 1
  ) +
  geom_node_label(
    aes(x = x, y = y, label = step),
    label.r = unit(0, units = "pt"),
    size = 3.5, lineheight = 1, fontface = "bold"
  ) +
  geom_edge_link(
    aes(
      start_cap = label_rect(node1.step, padding = margin(5, 5, 5, 5, "pt")),
      end_cap = label_rect(node2.step, padding = margin(5, 5, 5, 5, "pt"))
    ),
    arrow = arrow(angle = 30, length = unit(6, "pt"), type = "closed"),
    edge_linewidth = 0.25, color = "grey40"
  ) +
  scale_x_continuous(expand = expansion(0)) +
  scale_fill_manual(values = c("grey40", "grey80"), guide = "none") +
  expand_limits(y = 0.9) +
  theme_void()
```

Donor countries have a range of possible responses to recipient repression, including doing nothing and ignoring the violation, threatening to change aid allocations, withdrawing or suspending aid, or shifting aid to other projects or programs. In turn, recipient governments can then respond by doing nothing and ignoring the threat or suspension, change their behavior and cease the repression to improve their relationship with the donor country, or pressure the donor country to back down and withdraw their complaint. In this simplified decision tree, there are 12 possible combinations of outcomes,^[@CheesemanSwedlundOBrien-Udry:2024 describe additional stages in their tree, with donors responding to recipients' responses, creating dozens of possible pathways.] and each pathway is conditional on a host of donor-country and recipient-country characteristics and contexts, including geopolitical dynamics in the recipient country's region, domestic political institutions within the donor country, public opinion in the donor country, coordination between coalitions of donors, the type of aid, and the type of human rights violation, among others [@CheesemanSwedlundOBrien-Udry:2024]. Such complexity makes it difficult to find consistent treatment effects in research that explores the overall effect of repression on aid—that is, we cannot definitively say that human rights causes an overall reduction or increase in aid. Instead, newer research looks at donor responses in more specific circumstances, exploring the factors that lead to specific response pathways [@Corwin:2023].

### Donor responses to civil society laws

Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political oppression like respect for personal integrity, civil, and political rights [@Corwin:2023; @Neumayer:2003; @CheesemanSwedlundOBrien-Udry:2024], including arresting political activists, closing media outlets, suppressing protests, engaging in extrajudicial killings, and so on. The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored and understood.

Civil society plays a crucial role in the distribution and implementation of international aid. Governmental aid agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries. TODO: Something about the amount of money that goes through CS, perhaps highlighting just the US or UK. Larger international organizations also rely on domestic NGOs to carry out humanitarian and development projects. TODO: Something about the amount of money that the World Bank or UNHCR sends to NGOs. Outsourcing the provision of aid to NGOs allows donors to avoid aid capture in weak institutional environments and can lead to more effective program implementation [@Dietrich:2013; @CruzLabonneWright:2024]. Given the importance of NGOs in aid delivery, we have reason to expect that donor responses to civil society crackdown abroad are treated differently than more violent forms of repression, and thus likely to follow different pathways in @fig-agency-response-tree. 

Donor countries have a range of possible responses when recipient countries engage in administrative crackdown. Recent research finds that national and multilateral aid agencies are responsive to anti-NGO laws, though often in inconsistent ways [@ChaudhryHeiss:2022a]. @ChristensenWeinstein:2013 show that aid from bilateral donors—or direct country-to-country aid—sees an average decrease of $25–\$50 million after recipient states pass laws that restrict the flow of foreign funds to NGOs. However, aid from international, multilateral agencies is not affected by these laws. Similarly @DupuyPrakash:2018 find that the adoption of a restrictive foreign funding law is associated with a 32% decline in aid from national, bilateral donors, but has no corresponding change in aid from international, multilateral donors. 

One key reason for the divergence in the responses of bilateral and multilateral donors is the degree to which these donors rely on recipient-country NGOs to implement programs. Donor governments and agencies have different experiences with NGOs and correspondingly prioritize the openness of civil society in recipient countries in different ways. Multilateral donors like the World Bank and OECD tend to channel money directly to recipient governments rather than domestic NGOs and are thus more insulated from the effects of local civil society repression, which likely explains the lack of formal response to anti-NGO laws [@ChristensenWeinstein:2013; @DupuyPrakash:2018]. There is more variation, however, in how bilateral donors work with domestic NGOs. For example, @AllenFlynn:2018 find that the domestic political environments of donor countries shape how those countries channel their aid abroad. Because they tend to focus more on poverty alleviation initiatives, states with left-leaning governments typically channel more aid through recipient-country NGOs. In contrast, right-leaning governments are typically interested in promoting the economic and political interests of their own states, and accordingly prefer to channel aid directly through recipient-country governments [@AllenFlynn:2018]. Human rights abuses in recipient countries will shape the responses of donor countries depending on the saliency of the threat to aid provision: political repression that threatens business interests and investments will be more salient to right-leaning governments, while anti-NGO civil society administrative crackdown will likely have a stronger effect on left-leaning governments' responses, since domestic NGOs are central to donor countries' aid distribution strategies.

The nature of civil society restrictions also influences how official donor agencies respond. In their initial work on anti-NGO restrictions, @ChristensenWeinstein:2013 cataloged civil society laws into three broader categories—barriers to funding, barriers to entry, and barriers to advocacy—but they, and others like @DupuyPrakash:2018, dedicated the bulk of their analysis to looking at the effects of barriers to funding. This focus on funding restrictions is warranted, as most authoritarian civil society legal restrictions have been aimed specifically at reducing or controlling foreign funding [@Heiss:2019; @Carothers:2015a; @CarothersBrechenmacher:2014]. Moreover, the relationship between foreign aid and foreign funding restrictions appears to be bidirectional, as states are more likely to pass restrictive foreign funding laws in response to increases in bilateral aid [@DupuyRonPrakash:2016], especially if that aid arrives during competitive elections or times of political protest or instability [@Chaudhry:2022; @Heiss:2017]. Foreign funding restrictions and aid are thus linked: increasing aid leads to more funding restrictions [@DupuyRonPrakash:2016], which in turn lead to reductions in aid [@DupuyPrakash:2018].

Though the majority of anti-NGO laws have targeted foreign funding, other laws seek to impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation [@Chaudhry:2022]. Donor countries respond differently to these types of oppression. @ChaudhryHeiss:2024 find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law [see also @ChaudhryHeiss:2022a]. @RightSpringmanWibbels:2024 find similar and more nuanced trends in donor responses: donors committed to political advocacy   (i.e. those that fund democracy and civil society promotion activities) reduce funding for advocacy programs by more than 70% in response to new restrictions, and the reduction in aid persists for several years.

Overall bilateral aid does not decrease significantly in response to new barriers to entry, but the distribution channels of that aid shift as donor countries channel more aid through domestic rather than foreign NGOs [@ChaudhryHeiss:2024]. USAID even explicitly encourages its program managers to rely more on domestic, in-country NGOs when facing a restrictive legal environment [@USAID:2014]. In addition to changing the amounts and recipients of foreign aid, donor agencies often adjust the focus of the programs they fund, redirecting funding away from programs related to politically sensitive issues like human rights, anti-corruption, and democratic development and towards tamer topics that are more compatible with recipient-country preferences, such as education, health, and humanitarian services [@Bush:2015; @CheesemanDesrosiers:2023].

These shifts in funding patterns in response to restrictive civil society laws have had measurable effects on NGOs in recipient countries. Samantha Power, current administrator of USAID, has pledged to send 25% of USAID funding to local NGOs over the next few years [@Power:2021], though practical obstacles remain. New local partners take longer to vet and train compared to established INGO partners. Turning to local organizations entails greater risk, as domestic NGOs can be easily shut down based on the whims of the government. Additionally,  communities often distrust local groups that accept Western aid, especially those focused on democracy assistance [@Barkan:2012]. Donor preferences have also changed. Over the past two decades, bilateral donors have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015]. This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals [@Herrold:2020].


## Private donors

As explored above, given the centrality of aid in domestic foreign policy decisions and economic development strategy, substantial research in development economics, political science, and international relations has looked at the motivations behind bilateral and multilateral foreign aid, especially in response to human rights abuses in recipient states. However, states and international organizations are not the only institutions that engage in international giving. Private individuals exercise power and influence in unique ways that differ from other non-state actors. Since 2000, foreign aid funding from philanthropic foundations and individual donors in the United States has more than doubled [@DesaiKharas:2018], with nearly \$30 billion donated to international affairs and development organizations in 2023 [@GivingUSA:2024]. This growth in private foreign aid has had a substantial effect on overall international  aid flows and often serves as a complement to official bilateral and multilateral aid, particularly in response to acute humanitarian crises [@DesaiKharas:2018]. 

Understanding the dynamics of supporting international causes is important as philanthropy towards such causes can be harder to motivate. Work in nonprofit and philanthropic studies has long explored the personal motivations of individual donors, including reputation, altruism, and psychological benefits [see @BekkersWiepking:2011 for an exploration of eight different motivating mechanisms], but less research—particularly in political science and international  relations—has been done regarding the organizational or institutional mechanisms that drive donor behavior. Donors are motivated by the broader political and institutional contexts of NGO programming, and individuals are less likely to donate to foreign nonprofits and prefer giving to local causes, since the result of their giving is more visible [@CasaleBaumann:2015; @Tremblay-BoirePrakash:2017; @Wiepking:2010].

Donor motivations influence how individuals respond when potential recipient organizations face legal trouble abroad. However, the exact menu of responses is more limited than what is available to bilateral and multilateral states, as seen previously in @fig-individual-response-tree, and far more understudied. @fig-individual-response-tree presents a preliminary framework for outlining possible individual donor responses to anti-NGO repression. When a potential recipient NGO faces legal crackdown in its host country, thus limiting its operations, interested individual donors can respond by (1) increasing their donations, as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior. @ChaudhryDotsonHeiss:2021 explore how different combinations of donor characteristics determine which response individuals are likely to take when NGOs face crackdown. They find that donors with longer experiences with the nonprofit sector and high levels of social trust—i.e. those who frequently volunteer, regularly donate to charity, and trust political institutions—are more likely to maintain their support for international NGOs that face criticism or crackdown abroad. 

```{r}
#| label: fig-individual-response-tree
#| fig-cap: Possible individual responses to civil society restrictions in recipient countries
#| include: true
#| fig-width: 6
#| fig-height: 2.8
#| out-width: 80%

nodes <- tribble(
  ~id, ~x, ~y, ~step,
  1,   0,  3,  "Restrict civil society",
  2,   1,  4,  "Do nothing",
  3,   1,  3,  "Increase\ndonations",
  4,   1,  2,  "Decrease\ndonations",
  5,   2,  3,  "?"
)

edges <- tibble(
  from = c(1, 1, 1, 2, 3, 4),
  to =   c(2, 3, 4, 5, 5, 5)
)

actors <- tribble(
  ~actor, ~label, ~x,
  "Recipient", "NGO host government\naction", 0,
  "Donor", "Individual donor\nresponse", 1,
  "NGO", "NGO\nresponse", 2
) |>
  mutate(
    xmin = x - 0.5,
    xmax = x + 0.5
  )

tidy_graph <- tbl_graph(nodes = nodes, edges = edges) |> 
  create_layout(layout = nodes)

ggraph(tidy_graph) +
  geom_rect(
    data = actors, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = 4.2, fill = actor), inherit.aes = FALSE,
    alpha = 0.25
  ) +
  geom_rect(
    data = actors, aes(xmin = xmin, xmax = xmax, ymin = 4.2, ymax = Inf, fill = actor), inherit.aes = FALSE,
    alpha = 0.5
  ) +
  geom_text(
    data = actors, aes(x = x, y = 4.4, label = label), inherit.aes = FALSE,
    size = 3.5,
    fontface = "bold", lineheight = 1
  ) +
  geom_node_label(
    aes(x = x, y = y, label = step),
    label.r = unit(0, units = "pt"),
    size = 3.5, lineheight = 1, fontface = "bold"
  ) +
  geom_edge_link(
    aes(
      start_cap = label_rect(node1.step, padding = margin(5, 5, 5, 5, "pt")),
      end_cap = label_rect(node2.step, padding = margin(5, 5, 5, 5, "pt"))
    ),
    arrow = arrow(angle = 30, length = unit(6, "pt"), type = "closed"),
    edge_linewidth = 0.25, color = "grey40"
  ) +
  scale_x_continuous(expand = expansion(0)) +
  scale_fill_manual(values = c("grey40", "grey5", "grey80"), guide = "none") +
  expand_limits(y = c(1.8, 4.5)) +
  theme_void()
```

Additional experimental evidence examines how donors react to the knowledge that potential recipient NGOs face legal difficulties. In a vignette-based factorial experiment, @ChaudhryHeiss:2021 find that foreign civil society restrictions serve as a heuristic when deciding to donate—learning about NGO repression increases individual generosity and causes potential donors to be willing to give substantially more to legally restricted nonprofits. This effect is strongest for privately-funded human rights-focused NGOs that face legal crackdown, as individuals feel that the organization deserves their support. Several participants justified their increased support because the experimental organization is "doing good work in countries where it is tough for groups like them to operate and they
need all the help they can get" [@ChaudhryHeiss:2021, 496]. Building on this work, @ChaudhryDotsonHeiss:2024 use a survey experiment to examine how organizational characteristics (e.g., issue area, funding sources) and practices (e.g., financial transparency) interact with structural concerns like legal crackdown abroad. They find that, all else equal, legal crackdown on its own causes donors to reduce their support for nonprofits—individuals are 5–10 percentage points less likely to donate to an NGO that is criticized or restricted, compared to an organization that enjoys a friendly relationship with its host government. However, this reduction in support can be offset by organizational characteristics. Financial transparency and accountability protect each increase the probability of donations by nine percentage points under the worst conditions of legal crackdown. Individual donors thus typically reduce their support when seeing legal crackdown, but change their response and increase their when seeing other signs that the organization follows best practices and is more deserving.

The question of how organizations respond to different donor reactions to legal crackdown remains unexplored (see the far right panel of @fig-individual-response-tree). To our knowledge, no work has been done regarding individual-level nonprofit fundraising strategies in the face of closing civic space, opening promising avenues for future research.

While private donations have become an important source of funding in the age of closing civi space and increased restrictions on official foreign funding, they are unlikely to full supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid. Testimony at a 2017 US Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017]. Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.


# References
